---
title: "DATA 624 Project 2"
author: "Andrew Bowen, Josh Forster, John Cruz"
date: "2024-04-25"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(caret)
library(corrplot)
library(mice)
```


## Exploratory Data Analysis

First, we'll read in our training and testing datasets. We've converted them via a basic R script and they are available on our [GitHub here](https://github.com/andrewbowen19/data624Presentation/tree/main/data)


```{r}
# Read in training and testing datasets
train <- read.csv("https://raw.githubusercontent.com/andrewbowen19/data624Presentation/main/data/train.csv")
test <- read.csv("https://raw.githubusercontent.com/andrewbowen19/data624Presentation/main/data/test.csv")
```


We can take a brief look at our training data summary.
```{r}
summary(train)
```


We can take a look at the distribution of our outcome variable (`PH`)
```{r}
# Plot pH level distribution: our output variable
ggplot(train, aes(x=PH)) + geom_histogram()
```


### Data Wrangling


```{r}
# Convert Brand.Code to a factor as it's categorical
train$Brand.Code <- as.factor(train$Brand.Code)
```


## Imputation
Our dataset is missing values
```{r message=FALSE, warning=FALSE}
# Impute using the mice package
imputed <- mice(train, meth="pmm")

# Set the training set to be imputed values
train <- imputed$imp
```




## Feature Selection

Next, we'll look at

```{r}

correlations <- cor(as.matrix(train %>% dplyr::select(-c("Brand.Code"))))

corrplot(correlations)
```




## Modeling


### Multiple Linear Regression
First, we'll fit a basic multiple linear regression model to our data

## Model Evalutation




